# Measures of Occurrence

<!-- 
Hidden comments placeholder
---------------------------

To preview:
bookdown::preview_chapter("chapters/10_part_intro_epi/03_measures_of_association.Rmd")

Copy and paste:
👆**Here's what we did above:**

-->

<p class="under-construction"> `r fontawesome::fa("hammer", fill = "#000000", height="1em")` This chapter is under heavy development and may still undergo significant changes. </p>

# Useful resources

* Delete when first production draft is added to the book.

* [Modern Epidemiology, Chapter 4, Measures of Occurrence](https://wolterskluwer.vitalsource.com/reader/books/9781975166298/epubcfi/6/22)

* [Epi III OneDrive](https://uthtmc-my.sharepoint.com/personal/michael_b_cannell_uth_tmc_edu//_layouts/15/onedrive.aspx?login_hint=Michael%2EB%2ECannell%40uth%2Etmc%2Eedu&id=%2Fpersonal%2Fmichael%5Fb%5Fcannell%5Futh%5Ftmc%5Fedu%2FDocuments%2F03%20Teaching%2F01%20Courses%20%2D%20Current%2FEpi%20III)

# Notes

* Delete when first production draft is added to the book.

* Start with the [PPT](https://uthtmc-my.sharepoint.com/:p:/r/personal/michael_b_cannell_uth_tmc_edu/_layouts/15/Doc.aspx?sourcedoc=%7B42BFC14E-B01C-45CA-AC65-E107EBB7DDFB%7D&file=measures_of_occurrence.pptx&action=edit&mobileredirect=true) for this module. It looks like there is some useful stuff. 

* Also, add in stuff from the Modern Epi chapter on Measures of Effect and Measures of Occurrence (linked above). 

* Remember that there doesn't need to be a one-to-one correlation between Epi III modules and R4Epi Chapters. We can assign multiple chapters for a given module.

```{r echo=FALSE}
# Adding some helper functions to help with population plots.
# Set everything to echo=FALSE. The goal here isn't to teach learners how to make population plots. It is simply to use them as a visual aid for learning about populations and samples.
source("../../R/templates.R") # For colors
source("../../R/sample_ny.R") # To make sampling easier
source("../../R/make_pop.R")  # To make an example population
source("../../R/plot_pop.R")  # To make a population plot
```

```{r echo=FALSE}
# Create a population
set.seed(123)
pop <- make_pop(
  n_total = 1000,
  prob_exposed = 0.5,
  prob_outcome_exposed = 0.5,
  prob_outcome_unexposed = 0.1,
  arrange = TRUE
)
```

```{r echo=FALSE}
# Plot it
pop %>% 
  plot_pop(p_size = 3)
```

<!-- I would like to add counts to the facet ribbons -->

```{r echo=FALSE}
# Plot it with facets
pop %>% 
  plot_pop(p_size = 3) +
  facet_wrap(vars(e_o_f)) 
```

# Counts

```{r echo=FALSE}
set.seed(123)
# Create a 10x10 group of people
pop <- expand_grid(x = 1:10, y = 1:10)
pop$outcome <- sample_ny()
ggplot(pop, aes(x, y)) + 
  geom_point(size = 10, aes(color = outcome)) +
  scale_color_manual("Event", values = c(u_blue, u_orange)) +
  theme(
    panel.background = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

# Print for PowerPoint
ggsave("/users/bradcannell/Desktop/count_plot.png", width = 7.29, height = 4.51)
```

```{r echo=FALSE}
set.seed(123)
# Create a 10x10 group of people
pop <- expand_grid(x = 1:10, y = 1:10)
pop$outcome <- sample_ny()
# Align colors
pop %>% 
  arrange(desc(outcome)) %>% 
  # Renumber the grid
  mutate(
    x = rep(1:10, each = 10),
    y = rep(1:10, 10)
  ) %>% 
ggplot(aes(x, y)) + 
  geom_point(size = 10, aes(color = outcome)) +
  scale_color_manual("Event", values = c(u_blue, u_orange)) +
  theme(
    panel.background = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

# Print for PowerPoint
ggsave("/users/bradcannell/Desktop/count_plot_arranged.png", width = 7.29, height = 4.51)
```

```{r echo=FALSE}
# Try adding numbers
set.seed(123)
# Create a 10x10 group of people
pop <- expand_grid(x = 1:10, y = 1:10)
pop$outcome <- sample_ny()
# Align colors
pop %>% 
  arrange(desc(outcome)) %>% 
  # Renumber the grid
  mutate(
    x = rep(1:10, each = 10),
    y = rep(1:10, 10)
  ) %>% 
  # Create id numbers. The group_by is to make them appear in the order
  # we would typically count them in on the plot.
  mutate(id = row_number()) %>% 
  group_by(x) %>%
  mutate(id = rev(id)) %>% 
  ungroup() %>% 
ggplot(aes(x, y)) + 
  geom_point(size = 10, aes(color = outcome)) +
  geom_text(aes(label = id), color = "white") +
  scale_color_manual("Event", values = c(u_blue, u_orange)) +
  theme(
    panel.background = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

# Print for PowerPoint
ggsave("/users/bradcannell/Desktop/count_plot_arranged_numbered.png", width = 7.29, height = 4.51)
```


```{r eval=FALSE, echo=FALSE}
# For PowerPoint
table(pop$outcome)
```




