# Populations

<!-- 
Hidden comments placeholder
---------------------------

To preview:
bookdown::preview_chapter("chapters/10_part_intro_epi/03_measures_of_association.Rmd")

Copy and paste:
👆**Here's what we did above:**

-->

<p class="under-construction"> `r fontawesome::fa("hammer", fill = "#000000", height="1em")` This chapter is under heavy development and may still undergo significant changes. </p>

# Useful resources

* [Modern Epidemiology, Chapter 4, Measures of Occurrence](https://wolterskluwer.vitalsource.com/reader/books/9781975166298/epubcfi/6/22)

* [Epi III OneDrive](https://uthtmc-my.sharepoint.com/personal/michael_b_cannell_uth_tmc_edu//_layouts/15/onedrive.aspx?login_hint=Michael%2EB%2ECannell%40uth%2Etmc%2Eedu&id=%2Fpersonal%2Fmichael%5Fb%5Fcannell%5Futh%5Ftmc%5Fedu%2FDocuments%2F03%20Teaching%2F01%20Courses%20%2D%20Current%2FEpi%20III)

# Notes

* Start with the [PPT](https://uthtmc-my.sharepoint.com/:p:/r/personal/michael_b_cannell_uth_tmc_edu/_layouts/15/Doc.aspx?sourcedoc=%7B42BFC14E-B01C-45CA-AC65-E107EBB7DDFB%7D&file=measures_of_occurrence.pptx&action=edit&mobileredirect=true) for the Measures of Occurence module. It looks like there is some useful stuff. 

* Also, add in stuff from the Modern Epi chapter on Measures of Effect and Measures of Occurrence (linked above). 

* Remember that there doesn't need to be a one-to-one correlation between Epi III modules and R4Epi Chapters. We can assign multiple chapters for a given module.

```{r echo=FALSE}
# Libraries I'm loading to make the chapter graphics work. Hidden from readers.
# Chunk below contains packages that readers should load to follow along.
library(tidyr)
library(ggplot2)
library(templates)

# Load color template
data("my_colors")
```

```{r}
library(dplyr, warn.conflicts = FALSE)
```


# Populations

<!--
Adding some helper functions to help with population plots.
Set everything to echo=FALSE. The goal here isn't to teach learners how to make
population plots. It is simply to use them as a visual aid for learning 
about populations and samples.
-->

```{r echo=FALSE}
# Helper function for sampling No and Yes - Simplified
sample_ny <- function(n = 100, prob_y = 0.5) {
  sample(c("No", "Yes"), n, TRUE, c(1 - prob_y, prob_y))
}

# For testing
# sample_ny(prob_y = 0.1)
```


```{r echo=FALSE}
# Create a pop of size x with e prop exposed and o prop with outcome
make_pop <- function(n_total = 100, 
                     prob_exposed, 
                     prob_outcome_exposed,
                     prob_outcome_unexposed,
                     arrange = FALSE) {
  # Figure out the smallest integer that will be at least size
  # n_total when multiplied by 2. The idea is to figure out the dimensions
  # for the closest thing I can get to a square given n_total
  n_sqrt <- sqrt(n_total)
  n_sqrt_ceiling <- ceiling(n_sqrt)
  drop <- n_sqrt_ceiling^2 - n_total
  
  # Make coordinates for grid of points
  pop <- expand_grid(
    x = seq(n_sqrt_ceiling),
    y = seq(n_sqrt_ceiling)
  )
  
  # Drop of n_sqrt is uneven. Drop from bottom right corner.
  # High x, low y.
  pop <- pop %>% 
    arrange(desc(y)) %>% 
    slice(1:(n() - drop)) %>% 
    arrange(x, y)
  
  # I still want y to be base 1
  pop$y <- pop$y + (1 - min(pop$y))
  
  # Add exposed and unexposed
  # Helper function for sampling No and Yes - Simplified
  sample_ny <- function(n = 100, prob_y = 0.5) {
    sample(c("No", "Yes"), n, TRUE, c(1 - prob_y, prob_y))
  }
  
  # Add exposed
  pop$exposed <- sample_ny(n = n_total, prob_y = prob_exposed)
  
  # Add outcome
  pop$outcome <- NA_character_
  n_exp_y <- sum(pop$exposed == "Yes")
  n_exp_n <- sum(pop$exposed == "No")
  pop$outcome[pop$exposed == "Yes"] <- sample_ny(n_exp_y, prob_y = prob_outcome_exposed)
  pop$outcome[pop$exposed == "No"] <- sample_ny(n_exp_n, prob_y = prob_outcome_unexposed)
  
  # Add exposure-outcome group columns
  # To combine the legend, we need to have a single variable with exposure 
  # and outcome information.
  pop <- pop %>% 
    mutate(
      e_o = case_when(
        exposed == "Yes" & outcome == "Yes" ~ "a",
        exposed == "Yes" & outcome == "No" ~ "b",
        exposed == "No" & outcome == "Yes" ~ "c",
        exposed == "No" & outcome == "No" ~ "d"
      ),
      e_o_f = factor(
        e_o, c("a", "b", "c", "d"),
        c(
          "Exposed - Outcome", "Exposed - No Outcome", 
          "Not exposed - Outcome", "Not exposed - No outcome"
        )
      )
    )
  
  # Arrange 
  # Sometimes I want the exposed-unexposed to be haphazardly spread around the 
  # plot. Sometimes, I want the exposed next to the exposed and the unexposed 
  # next to the unexposed.
  if (arrange) {
    # Separate x and y from the rest of the data before arranging
    x_y <- select(pop, x, y)
    pop <- pop %>% 
      select(-x, -y) %>% 
      arrange(desc(exposed), desc(outcome))
    # Add x and y back
    pop <- bind_cols(x_y, pop)
  }
  
  # Return tibble
  pop
}

# For testing
# set.seed(123)
# make_pop(
#   n_total = 100,
#   prob_exposed = 0.5,
#   prob_outcome_exposed = 0.5,
#   prob_outcome_unexposed = 0.1,
#   arrange = FALSE
# )
```

```{r echo=FALSE}
# Pull orange and blue colors from templates package
u_orange <- filter(my_colors, description == "University Orange") %>% pull(hex)
u_blue <- filter(my_colors, description == "University Blue") %>% pull(hex)
```

```{r echo=FALSE}
# Create a plot helper function
plot_pop <- function(.data, p_size = 5) {
  
  # Store shape codes
  cir <- 16
  tri <- 17
  
  # Create plot
  p <- ggplot(.data, aes(x, y, color = e_o_f, shape = e_o_f)) + 
    geom_point(size = p_size) +
    scale_color_manual("Exposure-Outcome", values = c(u_orange, u_orange, u_blue, u_blue)) +
    scale_shape_manual("Exposure-Outcome", values = c(tri, cir, tri, cir)) +
    theme(
      panel.background = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank()
    )
  
  # Return plot object
  p
}

# For testing
# pop_plot(pop, 5)
```

```{r echo=FALSE}
# Create a population
set.seed(123)
pop <- make_pop(
  n_total = 1000,
  prob_exposed = 0.5,
  prob_outcome_exposed = 0.5,
  prob_outcome_unexposed = 0.1,
  arrange = TRUE
)
```

```{r echo=FALSE}
# Plot it
pop %>% 
  plot_pop(p_size = 3)
```

<!-- I would like to add counts to the facet ribbons -->

```{r echo=FALSE}
# Plot it with facets
pop %>% 
  plot_pop(p_size = 3) +
  facet_wrap(vars(e_o_f)) 
```