# Using Git and GitHub

<!--
Hidden comments placeholder
-->

<p class="under-construction"> `r fontawesome::fa("hammer", fill = "#000000", height="1em")` This chapter is under heavy development and may still undergo significant changes.</p>

```{r echo=FALSE}
# knitr::include_graphics("https://picsum.photos/200/300")
```

```{r message=FALSE}
# library(dplyr, warn.conflicts = FALSE)
```

There are at least three overarching reasons why you should learn to use Git and GitHub as part of your workflow when your projects include data and/or coding:

1. Versioning   
2. Collaborating    
3. Preservation    

We’ll elaborate on what each of these are next. Then, we will introduce you to Git and GitHub and explain why they are some of the best tools currently available to help you with versioning and collaborating. We’ll go ahead and warn you now — this chapter is long because Git and GitHub can be hard to wrap your mind around at first. In fact, using Git and GitHub still frequently causes us confusion and frustration. However, we still believe that the payoff is ultimately worth the upfront investment of time and frustration. Additionally, we will do our best to make this introduction as gentle, comprehensible, and practically applicable as possible. 
