# Introduction to data management

Way back in the [Getting Started](#managing-data) chapter, I told you that managing data includes all the things you may have to do to your data to get it ready for analysis. I also talked about the 80/20 "rule." The basic idea of the 80/20 rule is that data management is where you will spend the majority of your time and effort when you are involved in just about any project that makes use of data. Unfortunately, I can't cover strategies for overcoming every single data management challenge that you will encounter in epidemiology. However, in this part of the book, I will try to give you a foundation in some of the most common data management tasks that you will encounter. I will also try to point you towards some of the best tools and resources for data management that the R community has to offer.

## Multiple paradigms for data management in R

Before moving on to providing you with examples of how to accomplish specific data management tasks, I think this is the right point in the book to touch on a couple of high-level concepts that we have more or less ignored thus far.

R is pretty unique among the major statistical programming applications used in epidemiology in many ways. Among them is that R has multiple paradigms for data management. That's what I'm calling them anyway. What I mean by that is that there are 3 primary packages that the vast majority of R users use for data management. They are  base R, `data.table`, and `dplyr`. There is a tremendous amount of overlap in the data management tasks you can perform with base R, `data.table`, and `dplyr`, but the syntax for each is very different. As are the relative strengths and weaknesses.

In this book, we will primarily use the `dplyr` paradigm for data management. We will do so because I believe in using the best tool to get the job done. Currently, I believe that the best tool for managing data in R is usually `dplyr`, and especially when you are new to R. However, there will be cases where I will show you how to use base R to accomplish a task. Where I do this, it's because I think that base R is the best tool for the job or because I think you are very likely to see base R way used when you go looking for help with a related data management challenge and I don't want you to be totally clueless about what you're looking at. 

As of this writing, I've decided not to specifically discuss using the `data.table` package for data management. I think the `data.table` package is a great package, and I use it when I think it's the best tool for the job. However, I think the confusion caused by introducing `data.table` in this text aimed primarily at inexperienced R users would cause more problems than it would solve. The last thing I'll say about `data.table` for now is that you may want to consider learning more about `data.table` if you routinely work with very large data sets (e.g., millions of rows). For reasons that are beyond the scope of this book, `data.table` is currently much faster than `dplyr`. However, for most of the work I do, and all of what we will do in this book, the time difference will be imperceptible to you. Literally milliseconds.

## The dplyr package

At this point in the book, you've already been exposed to several of the most important functions in the `dplyr` package. You saw the `filter()` function in the [Speaking R's language](#packages) chapter, the `mutate()` function in the [chapter on exporting data](#coerce-a-numeric-variable), and the `summarise()` function all over the descriptive analysis part of the book. However, I mostly glossed over the details at those points. In this section, I want to dive just a tiny bit deeper into how the `dplyr` functions work -- but not too deep.

### The dplyr verbs

The `dplyr` package includes five main functions for managing data: `mutate()`, `select()`, `filter()`, `arrange()`, and `summarise()`. [These five functions are often referred to as the dplyr verbs](https://dplyr.tidyverse.org/). And, the first two arguments to all five of these functions are `.data` and `...`. Let's go ahead and discuss those two arguments a little bit more.

<p class="note"> 🗒**Side Note:** I don't want to give you the impression that `dplyr` only contains 5 functions. In fact, `dplyr` contains many functions, and they are all designed to work together in a very intentional way.</p>

### The .data argument

I first introduced you to data frames in the [Let's get programming](#data-frames) chapter and we've been using them as our primary structure for storing and analyzing data in ever since. The R language allows for other data structures (e.g., vectors, lists, and matrices), but data frames are the most commonly used data structure for most of the kinds of things we do in epidemiology. Thankfully, the `dplyr` package is designed specifically to help people like you and I manage data stored in _data frames_. Therefore, `dplyr` verbs always receive a data frame as an input and return a data frame as an output. Specifically, the value passed to the `.data` argument must always be a data frame, and you will get an error if you attempt to pass any other data structure to the `.data` argument. For example:

```{r}
# No problem
df <- tibble(
  id = c(1, 2, 3),
  x  = c(0, 1, 0)
)

df %>% 
  filter(x == 0)
```

```{r error=TRUE}
# Problem
l <- list(
  id = c(1, 2, 3),
  x  = c(0, 1, 0)
) 

l %>% 
  filter(x == 0)
```

### The ... argument

The second value passed to all of the `dplyr` verbs is the `...` argument. If you are new to R, this probably seems like a really weird argument. And, it kind of is! But, it's also really useful. The `...` argument (pronounced "dot dot dot") has special meaning in the R language. This is true for all functions that use the `...` argument -- not just `dplyr` verbs. The `...` argument can be used to pass multiple arguments to a function without knowing exactly what those arguments will look like ahead of time -- including entire expressions. For example:

```{r}
df %>% 
  filter(x == 0)
```

Above we passed a data frame to the `.data` argument of the `filter()` function. The second value we passed to the `filter()` function was `x == 1`. Think about it, `x` is an object (i.e. a column in the data frame), `==` is a function (remember that operators are functions in R), and `0` is a value. Together, they form an expression (`x == 0`) that tells R to perform a relatively complex operation -- compare every value of `x` to the value `0` and tell me if they are the same. If you are new to programming, this may not seem like any big deal, but it's really handy to be able to pass that much information to a single argument of a single function. 

If this is all really confusing to you, don't get too hung up on it right now. The `...` argument is an important component of the R language, but it isn't important that you fully understand it in order to use R. If nothing else, just know that that the `...` is the second argument to all the `dplyr` verbs, and it is generally where you will tell R what you want to _do_ to the columns of your data frame (i.e., keep them, drop them, create them, sort them, etc.).

### Non-standard evaluation

A final little peculiarity about the `tidyverse` packages -- `dplyr` being one of them -- that I want to discuss in this chapter is something called  **non-standard evaluation**. _How_ non-standard evaluation works really isn't that important for us. If I'm being honest, I don't even fully understand how it works "under the hood." But, it is one of the big advantages of using `dplyr`, and therefore worth mentioning. Do you remember the [section in the Let's get programming chapter on common errors](#some-common-errors)? In that section I wrote about how a vector that lives in the global environment is a different thing to R than a vector that lives as a column in a data frame in the global environment. So, `weight` and `class$weight` are different things, and if you want to access the weight values in `class$weight` then you have to make sure and write the whole thing out. But, have you noticed that we don't have to do that in `dplyr` verbs? For example:

```{r}
df %>% 
  filter(df$x == 0)
```

In the example above we wrote out the column name using dollar sign notation. But, we don't _have_ to:

```{r}
df %>% 
  filter(x == 0)
```

When we don't tell a `dplyr` verb exactly which data frame a column lives in, then the `dplyr` verb will assume it lives in the data frame that is passed to the `.data` argument. This is really handy for at least two reasons:

* It reduces the amount of typing we have to do when we write our code. 👏

* It makes it easier to glance at our code and see what it's doing. Without all the data frame names and dollar signs strewn about our code, it's much easier to see what the code is actually doing. 

Overall, non-standard evaluation is a great thing -- at least in my opinion. However, it will present some challenges that we will have to overcome if we plan to use `dplyr` verbs inside of functions and loops. Don't worry, we'll come back to this topic later in the book.

Now that you (hopefully) have a better general understanding of the `dplyr` verbs, let's go take a look at how to use them for data management.
